---
title: "Assignment 3: Data Exploration"
author: "Kayla Emerson"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setup, message = FALSE}
#load necessary packages
library(tidyverse)
library(lubridate)
library(here)

#check current working directory
getwd()

#upload ECOTOX neonicotinoid dataset
neonics <- read.csv(
  file = here('Data','Raw','ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = T
)

#upload litter and debris dataset 
litter <- read.csv(
  file = here('Data','Raw','NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = T
)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We are interested in the ecotoxicology of neonicotinoids on insects because they could leave lasting impacts on certain insect species, including pollinators, which are very important for several plants and crops. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Studying litter and woody debris that falls on the forest ground can give us more insight as to how that ecosystem stores and cycles nutrients and carbon because litter is the first step in decomposition. It can also tell us more about the biodiversity and the health of the ecosystem. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter was collected into elevated PVC pipes and ground traps
 2.Ground traps were sampled once per year, elevated traps were sampled more frequently 
 3. Mass data for samples were obtained to an accuracy of 0.01 grams

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dimensions of neonics}
#get dimensions of neonics 

dim(neonics)
length(neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Summary of neonics effects}
#get summary of the "effect" column of neonics
summary(neonics$Effect) 

#sort effect column 
sort(summary(neonics$Effect))
```

> Answer: Behavior, mortality, and population are the most common effects that are studied. Hormones, histology, and physiology are the least common effects studied. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Summary of neonics species}
#get summary of the studied species
summary(neonics$Species.Common.Name) 

#get top 6 species using maxsum
summary(neonics$Species.Common.Name, maxsum = 6)

```

> Answer: The six most commonly studied species in the dataset are all pollinators - bees and wasps. These species might be of interest over other insects because they help to pollinate other plants and are essential to ecosystems all over the globe. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Class of concentrations}
#get the class of "Conc.1..Author" 
class(neonics$Conc.1..Author.)

```

> Answer: The class of "Conc.1..Author" column in the dataset is factor because some values were labeled as "NR" or included "/" and we are unsure what that means in terms of numbers. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Graphical exploration}
#generate a plot of the number of studies conducted by pub. year using geom_freqpoly
ggplot(neonics) +
geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Make graph better}
class(neonics$Test.Location)

#Add test location as an additional variable, and categorize by color
ggplot(neonics) +
geom_freqpoly(aes(x = Publication.Year, color = Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are "Field natural" and "Lab." These are the most common over time but way more so since 1990. After 2010, "Lab" was by far the most common test location.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Bar graph of Endpoint counts}
#Create a bar graph of Endpoint counts
ggplot(data = neonics, aes(x = Endpoint)) +
geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: The two most common end points are LOEL and NOEL. LEOL stands for  Lowest-observable-effect-level which means it was the lowest dose (concentration)
producing effects that were significantly different (as reported
by authors) from responses of controls. NOEL stands for 
No-observable-effect-level which means it's the highest dose (concentration) 
producing effects not significantly different from responses of
controls. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Explore Litter data}
#get the class of collectDate
class(litter$collectDate)

#change collectDate from factor to date
litter$collectDate = ymd(litter$collectDate)

#confirm class change 
class(litter$collectDate)
#amazing, moving on

#use unique function to determine which dates litter was sampled
unique(litter$collectDate)
#only twice...
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Litter data exploration plots}
#determine different plots sampled using unique
unique(litter$plotID)

#determine different plots sampled using summary
summary(litter$plotID)
```

> Answer: The unique function spits out the different plot IDs as a list of levels and then says "12 level." The summary function displays each unique plot ID and how the count for each plot ID. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Bar graph of functionalGroup}
#create bar graph of functionalGroup counts
ggplot(data = litter, aes(x = functionalGroup)) +
geom_bar()

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Box plot and Violin plot}
#create boxplot of dryMass by functionalGroup
ggplot(litter) +
geom_boxplot(aes(x = functionalGroup,y = dryMass,))


#create violin plot of dryMass by functionalGroup
ggplot(litter) +
geom_violin(aes(x = functionalGroup,y = dryMass,))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is a more effective visualization option than the violin plot because it mainly displays the "center" of the data by displaying the quartiles and mean, and the outliers are not as distracting as they are in the violin plot. There appears to be several outliers, so the violin plot does not do a good job at summarizing the data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Mixed, Needeles, and Twigs/branches tend to have the highest biomass at the sites, according to the boxplot. 
